HO unification from object language to meta language

### 2024-03-15 paper/main.tex
w

\noindent
Since \elpiIn{Pm} is an higher-order unification variable,
of type \elpiIn{tm -> tm},
with \elpiIn{x}
in its scope, the unification problem \ref{problem:pa}
admits one solution:

@@

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify with substitution
$\sigma$, we write $\sigma t$ for the application of the substitution to
$t$, $\sigma \subseteq \sigma'$ when $\sigma$ is more general than $\sigma'$,
and we assume that the unification of our meta language is correct:

%
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}
\end{gather}

\newcommand{\C}[3]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3)}
\newcommand{\D}[3]{\ensuremath{\langle #1,#2 \rangle^{-1}\mapsto #3}}

We illustrate a compilation $\C{s}{t}{l}$ that
maps a term $s$ in \Fo{} to a term $t$ in \Ho and a list of links $l$.
The links connect unification variables in \Ho with variables
in \Fo{} and are used to decompile the assignment,
$\D{\sigma}{l}{\rho}$.

We represent a logic program run for a goal g as
a list steps $p$ of length $\mathcal{N}$ each made of a unification problem
between terms $\mathcal{S}_{p_l}$ and $\mathcal{S}_{p_r}$
taken from the set of all terms $\mathcal{S}$.
If a rule is used
twice in a run we make two copies of the terms (the predicates)
occurring in the head or premises. So a run is the conjunction of all
unification problems, where $\rho_0$ is the empty substitution
%
$$
\bigwedge_{p = 1 \ldots \mathcal{N}}
\underbrace{
  \rho_{(p-1)}\mathcal{S}_{p_l} \Uo \rho_{(p-1)}\mathcal{S}_{p_l} \mapsto \rho_p
}_\mathrm{step}
$$

We simulate this run as where $\sigma_0$ is the empty substitution
%
\newcommand{\solve}{\ensuremath{\mathrm{check}}}
\begin{gather*}
\mathcal{T} \times \mathbb{L} = \{ (t_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{l_j} \}\\
\bigwedge_{p = 1 \ldots \mathcal{N}}
 \underbrace{
  \sigma_{(p-1)}\mathcal{T}_{p_l} \Ue \sigma_{(p-1)}\mathcal{T}_{p_r} \mapsto \sigma_p' \land
  \solve~(\mathbb{L},\sigma_p')\mapsto \sigma_p
 }_\mathrm{step'}\\
\D{\sigma_{\mathcal{N}}}{\mathbb{L}}{\rho_{\mathcal{N}}}
\end{gather*}

@@

If we focus on a single unification problem, that is
$\mathcal{S} = \{ s_1, s_2 \}$ and 
$\C{s_1}{t_1}{l_1}$ and $\C{s_2}{t_2}{l_2}$ we write
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \stackrel{def}{=}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \solve~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land
    \D{\sigma''}{\{l_1,l_2\}}{\rho}
\end{array}
$$
\noindent
We write $s \in \llambda$ if all unif variables in
$s$ are applied to distinct bound variables.

>>>>

\noindent
Since \elpiIn{Pm} is an higher-order unification variable
of type \elpiIn{tm -> tm},
with \elpiIn{x}
in its scope, the unification problem \ref{problem:pa}
admits one solution:

@@

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify with substitution
$\sigma$, we write $\sigma t$ for the application of the substitution to
$t$, $\sigma \subseteq \sigma'$ when $\sigma$ is more general than $\sigma'$,
and we assume that the unification of our meta language is correct:
%
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}
\end{gather}

\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[3]{\ensuremath{\langle #1,#2 \rangle^{-1}\mapsto #3}}

We illustrate a compilation $\C{s}{t}{m}{l}$ that
maps a term $s$ in \Fo{} to a term $t$ in \Ho, a variable mapping $m$ and
list of links $l$.
The variable map connects unification variables in \Ho with variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{\rho}$. Links represent problematic sub-terms which
are linked to the unification variable that stands in their place in the
compiled term. These links are checked for or progress XXX improve....

We represent a logic program run for a goal g as
a list steps $p$ of length $\mathcal{N}$ each made of a unification problem
between terms $\mathcal{S}_{p_l}$ and $\mathcal{S}_{p_r}$
taken from the set of all terms $\mathcal{S}$.
If a rule is used
twice in a run we make two copies of the terms (the predicates)
occurring in the head or premises. So a run is the conjunction of all
unification problems, where $\rho_0$ is the empty substitution
%
\newcommand{\solve}{\ensuremath{\mathrm{check}}}
\newcommand{\step}[4]{\ensuremath{\mathrm{step}(#1,#2,#3) \mapsto #4}}
\newcommand{\stepD}[4]{\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4}}
\newcommand{\runD}[2]{\ensuremath{\bigwedge_{p = 1 \ldots #2} \step{#1}{p}{\rho_{(p-1)}}{\rho_{p}}}}
\newcommand{\run}[2]{\ensuremath{\mathrm{run}(#1,#2)}}
$$
\step{\mathcal{S}}{p}{\rho}{\rho'} := \stepD{\mathcal{S}}{p}{\rho}{\rho'}
$$
$$
\run{\mathcal{S}}{\mathcal{N}} := \runD{\mathcal{S}}{\mathcal{N}}
$$
$$
\bigwedge_{p = 1 \ldots \mathcal{N}}
\underbrace{
  \rho_{(p-1)}\mathcal{S}_{p_l} \Uo \rho_{(p-1)}\mathcal{S}_{p_l} \mapsto \rho_p
}_\mathrm{step}
$$

We simulate this run as where $\sigma_0$ is the empty substitution
%
\begin{gather*}
\mathcal{T} \times \mathbb{M} \times \mathbb{L} = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\\
\bigwedge_{p = 1 \ldots \mathcal{N}}
 \underbrace{
  \sigma_{(p-1)}\mathcal{T}_{p_l} \Ue \sigma_{(p-1)}\mathcal{T}_{p_r} \mapsto \sigma_p' \land
  \solve~(\mathbb{L},\sigma_p')\mapsto \sigma_p
 }_\mathrm{step'}\\
\D{\sigma_{\mathcal{N}}}{\mathbb{L}}{\rho_{\mathcal{N}}}
\end{gather*}

@@

If we focus on a single unification problem, that is
$\mathcal{S} = \{ s_1, s_2 \}$ and 
$\C{s_1}{t_1}{m_1}{l_1}$ and $\C{s_2}{t_2}{m_2}{l_2}$ we write
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \stackrel{def}{=}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \solve~(\{l_1,l_2\},\sigma') \mapsto \sigma'' \land
    \D{\sigma''}{\{l_1,l_2\}}{\rho}
\end{array}
$$
\noindent
We write $s \in \llambda$ if all unif variables in
$s$ are applied to distinct bound variables.
